Are populations like a circuit? The relationship between isolation by distance and isolation by resistance

### 2018-08-09 isolation_by_coalescence.tex
a few more edits

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Then we define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
A similar argument using the Markov property
shows that $C$ satisfies the following equation:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
This can also be seen by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \otimes I + I \otimes G$,
where $I$ is the identity matrix and $\otimes$ is the Kronecker product.

@@

\paragraph{Constraints and uniqueness}
We have seen that to find the movement rates of a continuous-time Markov chain
given the coalescence or commute times
we must solve equations \eqref{eqn:C_matrix} or \eqref{eqn:R_from_H} and \eqref{G_from_H} for $G$,
and that these do not hae unique solutions.
The situation is worse when one needs to infer coalescence rates ($\gamma$) as well.
However, usually in applications the locations lie across geographical space,
so that many of the entries of $G$ can be assumed to be zero.
We can get an idea of this by simply counting equations and unknowns.
For concreteness, suppose that the spatial locations
are arranged in a square grid of $n$ locations,
so that each is connected to four others.
(Boundary locations will have fewer, but we omit this detail.)
Since movement rates in each direction can be different,
there are $4n$ free parameters, each corresponding to an offdiagonal entry of $G$.
Coalescence rates provide another $n$ parameters.

@@

\begin{figure}
\centering
     \includegraphics[scale=.8]{figs/4x4coalvcomH}
    \caption{Plots of the true values of $\comdist$ vs $H$ 
    for the uniform graph, symmetric graph, the asymmetric graph,
    and the biased asymmetric graph, 
    shown with the $y=x$ line.
    See Figure \ref{fig:4x4_grids} to see overall graph structure.}
    \label{fig:4x4coalvcomH}
\end{figure}

@@

\paragraph{Process noise}
Talk about Figure \ref{fig:h_val_comp}.

@@

\begin{figure}
\centering
     \includegraphics[scale=.8]{figs/post_dists_5x3b_1_new}
    \caption{Boxplots of the posterior distributions for $g$ 
    for the first replicate of the $5 \times 3$ landscape with barriers.
    Boxes for gene flow rates for edges that cross the barrier are colored in red.}
    \label{fig:post_dists_5x3b_1}
\end{figure}

@@

\subsection*{Continuous Landscape with Biased Migration}

The third situation that we will discuss is where migration is biased.
We will be using a square landscape as in the first case, 
but the offspring locations of offspring are not centered around the parent.
Instead, the mean of the distribution of the locations
is one tenth of a standand deviation (or 0.05\% of the edge length of the landscape) 
up and to the right of their parent's location.
This results in reverse time gene flow down and to the left 
as in the biased asymmetric graph in Section \ref{sec:biased_migration}.
In terms of population genetics, 
this is very similar to the case where a population is expanding into new territory 
since in both cases, 
offspring tend to be further from the edge of the expansion than their parents.
The difference is that, in our case, 
the population is constrained to be a bounded distance 
away from the edge of the ``expansion''.
In both situations, individuals at the edge may have disproportionately more offspring,
leading to low genetic diversity and strong drift.
This results in substantially more noise in the isolation by distance plots 
as seen in Figure \ref{fig:ibd_comp}. 
The horizontal lines in that figure are likely from comparing genomes of individuals from
two families that have recently increased in size and geographical spread, 
resulting in many pairs of genomes with similar genetic distance 
but varying geographic distance.

@@

\begin{figure}
\centering
 \includegraphics[scale=.34]{figs/tort_land5}
\caption{Left: Locations of the sampled tortoises.
    \plr{add mountain shadings and ocean; remove bits on the other side of the Colorado}
    Right: Sample maximum posterior estimates 
    of movement rates.
    \plr{can these be maximum posterior estimates instead?}
    } \label{fig:tort_land}
\end{figure}

\begin{figure}
\centering
 \includegraphics[width=0.8\textwidth]{figs/tort_post_g}
 \includegraphics[width=0.8\textwidth]{figs/tort_post_gam}
\caption{
Boxplots of the posterior distributions for 
\textbf{(top)} movement rates, and
\textbf{(bottom)} coalescence rates,
for the Mojave desert tortoise data set.
Parameters are labeled as in Figure \ref{fig:tort_land}.
\label{fig:tort_post}
\end{figure}

\begin{figure}
\centering
 \includegraphics[scale=.8]{figs/tort_h_comp}
\caption{Mean coalescence times calculated from the sample maximum likelihood model parameters 
plotted against the observed mean coalescence times calculated from the data, 
measured in the mean of pairwise $\pi$.
The mean absolute value relative differences between the two is 0.4\%.}
\label{fig:tort_h_comp}
\end{figure}

@@

resistance distance can mislead, especially in the presence of biased migration

coalescence time analogue is feasible but more computationally intensive

practical example (tortoises) shows strong asymmetry

but methods dealing with ill-posedness are necessary

other measures of genetic distance: what do people use in practice?

@@

\plr{from the thesis; merge in}
With these methods, we provide a framework for
inferring demographic parameters of 
populations living on a habitat spread over a two dimensional landscape 
by using present day location and genetic sequencing information.
We are able to do inference of reverse time gene flow and coalescence rates 
with the matrix of mean coalescence times
in cases where the matrix is noise-free, noisy, or incomplete, 
or when it made after discretization of a simulated or real landscape.
To do this, we assume that the reverse time movements of lineages are Markov
and the population is near stationarity.
Typically, there is more error and uncertainty in the posterior distributions of the parameters
when there is more noise, slower coalescence rate, more missing values,
and multiple parameters for coalescence rate.

While the primary focus is coalescence time based inference,
we also compare the results with that of resistance distance (i.e., commute time) based inference.
Coalescence based inference typically has lower error than commute time based inference,
along with the benefit that you can also infer coalescence rates. 
The commute time plus diversity estimate, $\comdist$, 
provides a good estimate of the coalescence times, $H$,
when gene flow rates are symmetric ($G_{ij} = G_{ji}$), 
but can be very wrong when they are asymmetric, 
including extreme discrepencies 
when there is an overall bias in the direction of the asymmetry across the landscape 
(Figure \ref{fig:4x4coalvcomH}).
When inference is performed on discrete landscapes 
with varying degrees of migrational asymmetry, 
coalescence time based inference is more accurate than commute time inference in all tested cases
(Figure \ref{fig:4x4box}).
Although performance of the commute time based method is reasonable 
when gene flow rates in the underlying model are symmetric, 
it becomes worse as the gene flow rates become more asymmetric 
to the point where there is little correlation at all with the true values 
when there is biased asymmetric migration.
On the other hand, commute time inference is much less computationally intensive, 
so much larger spatial systems can be used.
If the gene flow rates are symmetric, the higher resolution allowed by commute time inference
may outweigh the loss in accuracy.
The implication of these results for the body of literature using resistance distance methods
is that one may need to use caution when interpreting results.
For example, inferring high gene flow rates between two areas by using resistance distance
could potentially mean that those were areas were actually being fed by a common source.
Therefore, it may be worthwhile to apply a coarse resolution coalescence time based method 
alongside a finer resolution resistance distance method 
in order to check for evidence of asymmetric gene flow patterns.

\plr{pasted in: nonmarkov lineages}
Even though the population dynamics in forwards time are Markov, the dynamics of lineages traced back
in time (the coalescent process) may not be. For example, if fecundity is variable and population density
is low, two lineages near each other are more likely to share a recent common origin, so knowing how one
lineage moves gives you information about how the other lineage is likely to move. This effect becomes
negligible as population density increases [Barton et al., 2002]), so it should not be an issue if the offspring
of any one parent are typically interspersed with the offspring of many others. The methods I develop here
model the coalescent process as Markov, which should be a good approximation for many systems.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Finding $G$ from $R$}

The resistance distance is given by
$R = H + H^T$, so if we can modify $H$ by a skew-symmetric matrix,
we won't change $R$.
Can we do this and retain a valid $G$?
I.e., does $R$ uniquely specify $G$?
Let $Z + Z^T = 0$. 
Since $G$ is a continuous function of $H$, if all entries of $G$ are nonzero
we can find a small enough $\epsilon$ such that the $G'$ corresponding to $H' = H + \epsilon Z$
is also a valid generator matrix.
If $G$ has zero entries this may not be true.
To find a direction that we can change $H$ in that doesn't mess this up,
we could differentiate $dG/dH$.

@@

\begin{figure}
\centering
     \includegraphics[scale=.8]{figs/grid_5x3b_1}
    \caption{Medians of the posterior distributions of the inferred values of $g$ 
    for the first replicate of the $5 \times 3$ landscape with barriers.
    The gene flow rate across the barriers are small, 
    but there is no clear pattern amoung the rest of the values.}
    \label{fig:grid_5x3b_1}
\end{figure}

>>>>

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Then we define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
A similar argument using the Markov property
shows that $C$ satisfies the following equation:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
In practice, we solve this by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \otimes I + I \otimes G$,
where $I$ is the identity matrix and $\otimes$ is the Kronecker product.
\plr{say a bit more about how we solve this?}

@@

\paragraph{Constraints and uniqueness}
We have seen that to find the movement rates of a continuous-time Markov chain
given the coalescence or commute times
we must solve equations \eqref{eqn:C_matrix} or \eqref{eqn:R_from_H} and \eqref{eqn:G_from_H} for $G$,
and that these do not have unique solutions.
The situation is worse when one needs to infer coalescence rates ($\gamma$) as well.
However, usually in applications the locations lie across geographical space,
so that many of the entries of $G$ can be assumed to be zero.
We can get an idea of this by simply counting equations and unknowns.
For concreteness, suppose that the spatial locations
are arranged in a square grid of $n$ locations,
so that each is connected to four others.
(Boundary locations will have fewer, but we omit this detail.)
Since movement rates in each direction can be different,
there are $4n$ free parameters, each corresponding to an offdiagonal entry of $G$.
Coalescence rates provide another $n$ parameters.

@@

\begin{figure}
\centering
     \includegraphics[scale=.8]{figs/4x4coalvcomH}
    \caption{Plots of the true values of $\comdist$ vs $H$ 
    for the uniform graph, symmetric graph, the asymmetric graph,
    and the biased asymmetric graph, 
    shown with the $y=x$ line.
    \plr{TODO: make this a 2x2 grid instead of 4x1}
    See Figure \ref{fig:4x4_grids} to see overall graph structure.}
    \label{fig:4x4coalvcomH}
\end{figure}

@@

\paragraph{Process noise}
The mean coalescence times of lineages that we compute analytically
are exact for large, randomly mating populations,
but the word ``mean'' indicates an average across several levels of randomness:
and observed, empirical mean estimated using a sample of genotyeps
will differ slightly depending on both the individuals sampled (``sampling noise'')
and the stochasticity of the population history itself (``process noise'').
For more discussion of this latter contribution in other contexts, see
\citet{waples}, \citet{wakeley}, or \citet{ralph_empirical}.
In populations living across continuous geography,
both sources of noise can be substantially larger due to randomness in spatial locations.
A hint of this is visible in Figure \ref{fig:ind_locs_5x3b_1},
where we see that the distribution of individuals across the landscape
is fairly patchy.
This might make inference using a discretization of continuous space considerably more difficult.

To quantify the relative contributions of the different sources of noise,
we coompared mean genetic distances, calculated in the same way,
between nonoverlapping sets of samples from the same simulation,
and between different simulations.
These simulations were done using SLiM as described in the Methods
on a square, $4 \times 4$ landscape with no barriers.
We found that process noise was much larger than sampling noise --
as shown in Figure \ref{fig:h_val_comp},
discrepancies between simulations are about five times that seen
between different samples from the same simulation.
An implication of this is that uncertainties in genetic distance
estimated from the data will be (possibly large) underestimates,
a fact also pointed out for effective population size by \citet{waples}.

@@

\begin{figure}
\centering
     \includegraphics[height=0.3\textheight]{figs/grid_5x3b_1}
     \includegraphics[height=0.3\textheight]{figs/post_dists_5x3b_1_new}
    \caption{
    \textbf{(top)}
    Posterior medians of movement rates for a replicate of the continuous landscape
    with barriers depicted in Figure \ref{fig:ind_locs_5x3b_1}.
    \textbf{(bottom)}
    Boxplots of the same posterior distributions,
    colored red for gene flow rates of edges that cross the barrier.
    } \label{fig:post_dists_5x3b_1}
\end{figure}

@@

\paragraph{Continuous Landscape with Biased Migration}
The third situation that we will discuss is where migration is biased.
We will be using a square landscape as in the first case, 
but the offspring locations of offspring are not centered around the parent.
Instead, the mean of the distribution of the locations
is one tenth of a standand deviation (or 0.05\% of the edge length of the landscape) 
up and to the right of their parent's location.
This results in reverse time gene flow down and to the left 
as in the biased asymmetric graph in Section \ref{sec:biased_migration}.
In terms of population genetics, 
this is very similar to the case where a population is expanding into new territory 
since in both cases, 
offspring tend to be further from the edge of the expansion than their parents.
The difference is that, in our case, 
the population is constrained to be a bounded distance 
away from the edge of the ``expansion''.
In both situations, individuals at the edge may have disproportionately more offspring,
leading to low genetic diversity and strong drift.
This results in substantially more noise in the isolation by distance plots 
as seen in Figure \ref{fig:ibd_comp}. 
The horizontal lines in that figure are likely from comparing genomes of individuals from
two families that have recently increased in size and geographical spread, 
resulting in many pairs of genomes with similar genetic distance 
but varying geographic distance.

@@

\begin{figure}
\centering
 \includegraphics[width=0.45\textwidth]{figs/fancy_watershed_assignments}
 \includegraphics[width=0.45\textwidth]{figs/tort_graph_results}
\caption{Left: Locations of the sampled tortoises.
    \plr{add mountain shadings and ocean; remove bits on the other side of the Colorado}
    Right: Sample maximum posterior estimates 
    of movement rates.
    \plr{can these be maximum posterior estimates instead?}
    } \label{fig:tort_land}
\end{figure}

\begin{figure}
\centering
 \includegraphics[width=0.8\textwidth]{figs/tort_post_g}
 \includegraphics[width=0.8\textwidth]{figs/tort_post_gam}
\caption{
    Boxplots of the posterior distributions for 
    \textbf{(top)} movement rates, and
    \textbf{(bottom)} coalescence rates,
    for the Mojave desert tortoise data set.
    Parameters are labeled as in Figure \ref{fig:tort_land}.
    } \label{fig:tort_post}
\end{figure}

\begin{figure}
\centering
 \includegraphics[scale=.8]{figs/tort_h_comp}
\caption{
    Mean coalescence times calculated from the sample maximum likelihood model parameters 
    plotted against the observed mean coalescence times calculated from the tortoise data, 
    measured in the mean of pairwise $\pi$.
    The mean absolute value relative differences between the two is 0.4\%.
    } \label{fig:tort_h_comp}
\end{figure}

@@

With these methods, we provide a framework for
inferring demographic parameters of 
populations living on a habitat spread over a two dimensional landscape 
by using present day location and genetic sequencing information.
We are able to do inference of reverse time gene flow and coalescence rates 
with the matrix of mean coalescence times
in cases where the matrix is noise-free, noisy, or incomplete, 
or when it made after discretization of a simulated or real landscape.
To do this, we assume that the reverse time movements of lineages are Markov
and the population is near stationarity.
Typically, there is more error and uncertainty in the posterior distributions of the parameters
when there is more noise, slower coalescence rate, more missing values,
and multiple parameters for coalescence rate.

While the primary focus is coalescence time based inference,
we also compare the results with that of resistance distance (i.e., commute time) based inference.
Coalescence based inference typically has lower error than commute time based inference,
along with the benefit that you can also infer coalescence rates. 
The commute time plus diversity estimate, $\comdist$, 
provides a good estimate of the coalescence times, $C$,
when gene flow rates are symmetric ($G_{ij} = G_{ji}$), 
but can be very wrong when they are asymmetric, 
including extreme discrepencies 
when there is an overall bias in the direction of the asymmetry across the landscape 
(Figure \ref{fig:4x4coalvcomH}).
When inference is performed on discrete landscapes 
with varying degrees of migrational asymmetry, 
coalescence time based inference is more accurate than commute time inference in all tested cases
(Figure \ref{fig:4x4box}).
Although performance of the commute time based method is reasonable 
when gene flow rates in the underlying model are symmetric, 
it becomes worse as the gene flow rates become more asymmetric 
to the point where there is little correlation at all with the true values 
when there is biased asymmetric migration.
On the other hand, commute time inference is much less computationally intensive, 
so much larger spatial systems can be used.
If the gene flow rates are symmetric, the higher resolution allowed by commute time inference
may outweigh the loss in accuracy.
The implication of these results for the body of literature using resistance distance methods
is that one may need to use caution when interpreting results.
For example, inferring high gene flow rates between two areas by using resistance distance
could potentially mean that those were areas were actually being fed by a common source.
Therefore, it may be worthwhile to apply a coarse resolution coalescence time based method 
alongside a finer resolution resistance distance method 
in order to check for evidence of asymmetric gene flow patterns.

@@

\paragraph{The effect of history}
The models we discuss here assume that population sizes and migration rates
have been constant on the time scale given by the within-species coalescent time.
This is rarely true in practice \citep{avise,barton}.
However, geographic differences in mean relatedness
are established on a shorter time scale --
the time scale over which a lineage, moving randomly across the landscape,
``forgets'' where it started
(i.e., the mixing time of the Markov chain).
This is the time it takes for the uncertainty in lineage movement
to reach the scale of the species range.
For instance,
the width of the tortoise range is roughly 500km,
and if mean dispersal distance is 10km (a high estimate),
the standard deviation of ancestor location $t$ generations ago is $10\sqrt{t}$,
so the mixing time of tortoise lineages across the modern landscape,
not accounting for barriers, is of order 2500 generations,
or about 50,000 years.
The landscape of the Mojave has changed substantially over tihs time \citep{pleistocene_mojave},
which suggests that models incorporating change over time 
may be required to model modern tortoise diversity.
However, the effects of recent history are strongest,
so landscapes estimated assuming constant populations
may give a reasonable picture of the landscape averged over recent times.

\paragraph{Assumptions}
Modeling lineages as a Markov chain is nearly ubiquitous in population genetics today,
but may not be appropriate.
Even though the population dynamics in forwards time are Markov, the dynamics of lineages traced back
in time (the coalescent process) may not be. For example, if fecundity is variable and population density
is low, two lineages near each other are more likely to share a recent common origin, so knowing how one
lineage moves gives you information about how the other lineage is likely to move. This effect becomes
negligible as population density increases [Barton et al., 2002]), so it should not be an issue if the offspring
of any one parent are typically interspersed with the offspring of many others. The methods I develop here
model the coalescent process as Markov, which should be a good approximation for many systems.

\paragraph{other points:}
Practical example (tortoises) shows strong asymmetry.
Methods dealing with ill-posedness are necessary.
Other measures of genetic distance: what do people use in practice?

@@

@@
